---
title: "Essence of Calculus"
author: "Emily Schwenger"
date: "`r Sys.Date()`"
output: 
  rmdformats::readthedown:
      self_contained: true
      thumbnails: false
      lightbox: true
      gallery: false
      highlight: tango
      df_print: kable
      fig_height: 6.5
      fig_width: 10
      css: custom.css
      toc_depth: 3
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE) 
```

# Setup

## Load libraries

```{r}
suppressPackageStartupMessages( {
  library(RColorBrewer)
} )
```

## Color palettes

```{r}
pal <- brewer.pal(11, name="Spectral") 
cl1 <- "red4"
cl2 <- "peru"
```

# Fundamental theorem of calculus

If $f(x)$ is continuous over interval $[a,b]$, and $F(x)$ is defined as $F(x)=\int_a^x f(t)dt$, then $F'(x)=f(x)$ over $[a,b]$.

$dx=\Delta x \rightarrow 0$

$\dfrac{d}{dx} \int_a^xf(t)dt=f(x)$

$\int_a^b f(x)dx=F(B)-F(A)$ where $F(X)=\int f(x)$

# Derivatives

## General form of a derivative

$\dfrac{ds}{dt}(s)=\dfrac{s(t+dt)-s(t)}{dt}$ as $dt \rightarrow 0$, i.e. we are taking tinier and tinier steps to be able to approximate an "instantaneous rate". Posing the problem as a limit prevents us from stumbling into paradoxes due to the seemingly counterintuitive phrase "instantaneous rate" since a rate by definition requires a rate of change between two different points. 

## Rules

### Constant rule

$\dfrac{d}{dx}(c)=0$

### Constant multiple rule

$\dfrac{d}{dx}(cf(x))=cf'(x)$

### Power rule

$\dfrac{d}{dx}(x^n)=nx^{x-1}$

### Sum rule

$\dfrac{d}{dx}(f(x)+g(x))=f'(x)+g'(x)$

### Product rule

"Left d(Right) + Right d(Left)"

$\dfrac{d}{dx}(f(x)g(x))=f(x)g'(x) + f'(x)g(x)$

### Quotient rule

$\dfrac{d}{dx}[\dfrac{f(x)}{g(x)}]=\dfrac{g(x)f'(x)-f(x)g'(x)}{[g'(x)]^2}$

### Chain rule

$\dfrac{d}{dx}(g(h(x))=\dfrac{dg}{dh}\dfrac{dh}{dx}=\dfrac{dg}{dx}$

## Exponentials and Euler's number

Remember this fundamental rule of exponents...

$2^{a+b}=2^a2^b$

Let's say we have a population that is exponentially growing with the doubling function $M(t)=2^t$.

Then,

$\dfrac{dM}{dt}(t)=\dfrac{2^{t+dt}-2^t}{dt}=\dfrac{2^t2^{dt}-2^t}{dt}=2^t[\dfrac{2^{dt}-1}{dt}]$. Since $dt \rightarrow 0$, we can plug in smaller and smaller values of $dt$ into the second part of the equation $\dfrac{2^{dt}-1}{dt}$ and we will see that we converge at a constant...

```{r}
dt <- 0.01 
(2^dt-1)/dt
dt <- 1e-4 
(2^dt-1)/dt
dt <- 1e-6 
(2^dt-1)/dt
dt <- 1e-10 
log(2, exp(1))
( round( (2^dt-1)/dt, 5) )==round( log(2, exp(1)), 5 )

0.6931478*3
2^3

dt <- 0.01 
(8^dt-1)/dt
dt <- 1e-4 
(8^dt-1)/dt
dt <- 1e-6 
(8^dt-1)/dt
dt <- 1e-10 
(8^dt-1)/dt
```

So this is telling us that the derivative of $2^t$ is just itself times a constant, i.e. $\dfrac{d(2^t)}{dt}=c2^t$.

Q: Is there a base where that proportionality constant is 1? 
A: Yes! Euler's number

```{r}
e <- exp(1)
dt <- 0.01 
(e^dt-1)/dt
dt <- 1e-4 
(e^dt-1)/dt
dt <- 1e-6 
(e^dt-1)/dt
dt <- 1e-10 
(e^dt-1)/dt
```

This is what defines the number e. The fact that, when raised to an independent variable, it is a derivative of itself.

$\dfrac{d(e^x)}{dx}=e^x$, or, more generally...

$\dfrac{d(e^{cx})}{dx}=ce^{cx}$

It's a bit like asking "Why does pi happen to be the ratio of the circumference of a circle to its diameter?" This is circular, because this is what *defines* this value.

<mark >All exponential functions are proportional to their own derivative, but e alone is equal to its own derivative</mark>.

$2=e^{ln(2)}$ (by definition), i.e. `ln(x)` is the inverse of `e^x`.
$2^t=e^{ln(2)t}$

$\dfrac{2^x-1}{x}=ln(2)$ for small $x \rightarrow 0$.

SO, in conclusion, we have:

$\dfrac{d(2^x)}{dx}=2^xln(2)$

Try to write anything in exponential form with $e$ as the base to make the intuition easier!

$a^x=e^{cx}$. 

E.g. $\dfrac{d(\Delta T)}{dt}=-k \Delta T$ so $\Delta T(t)=e^{kt}$, i.e. the rate of change in temperature is proportional to the difference in temperature itself. This is a very natural concept.

## Implicit derivatives

Sometimes functions do not have separate independent and dependent variables and are rather on the same side of the equation, i.e. interdependent in a certain relation. 

### E.g. $x^2 + y^2 = 5^2$.

To calculate the derivative, you derive both sides of the equation, giving

$2xdx + 2ydy = 0$

$\dfrac{dy}{dx}=-\dfrac{x}{y}$

### E.g. $sin(x)y^2=x$

$sin(x)y^2=x$. Derive both sides using "Left d-right plus right d-left"...

$sin(x)2ydy+y^2cos(x)dx=dx$

$\dfrac{dy}{dx}=$

### E.g. $y=ln(x)$

Remember, $ln(x)$ is the inverse function of $e^x$.

$y=ln(x)$

$e^y=x$

$e^ydy=dx$

$\dfrac{dy}{dx}=\dfrac{1}{e^y}=\dfrac{1}{x}$.

# Limits

## Formal definition of derivatives

General form: $\dfrac{ds}{dt}(s)=\dfrac{s(t+dt)-s(t)}{dt}$ as $dt \rightarrow 0$

$\dfrac{df}{dx}(2)=\lim_{h \rightarrow 0} \dfrac{f(2+h)-f(2)}{h}$

## $(\epsilon, \delta)$ definition of limits

Glimpse into *real analysis*. Idea is that when a limit exists you can make the output range as small as you want, but when the limit does *not* exist, that *output* range can*not* get smaller than some particular value (e.g. for discontinuous functions) no matter how much you shrink the *input* range.

What it means for the limit to *exist* is that you can always find a range of inputs around our limiting input some distance **delta** from 0, such that any *input* within a distance *delta* of 0 corresponds to an output within a distance **epsilon** of 12 (or other number on right of equation).

## L'Hôpital's rule (irl Bernoulli's rule)

$lim_{x \rightarrow a}\dfrac{f(x)}{g(x)}=\dfrac{ \dfrac{df}{dx}(a)dx}{\dfrac{dg}{dx}(a)dx}=\dfrac{ \dfrac{df}{dx}(a)}{\dfrac{dg}{dx}(a)}$

### E.g. Dampened oscillation $\dfrac{sin\pi x}{x^2-1}$

At $x=(-1,1)$ denominator is 0.

Derivative of $sin(\pi x)$:

$d(sin(\pi x))=\pi cos(\pi x)dx$ and at $x=1$ we have 

$\pi cos(\pi)dx=-\pi dx$.

Similarly for $d(x^2-1)=2xdx=2dx$. Plugging it into the formal derivative equation, we have,

$lim_{x \rightarrow 1}\dfrac{sin\pi x}{x^2-1} \approx \dfrac{-\pi dx}{2dx}\approx \dfrac{-\pi }{2}$

```{r}
x <- seq(-2*pi, 2*pi, 0.1)
y <- sin(pi*x)/(x^2-1)

x_illeg <- c(-1.000000001, 1.000000001)
y_illeg <- sin(pi*x_illeg)/(x_illeg^2-1)
round(y_illeg[2], 6)==round(-pi/2, 6)

plot(x, y, type="l", col=pal[1], lwd=3)
abline(h=0, v=0, lwd=2)
points(rep(x_illeg,2), c(y_illeg,0,0), pch=21, bg="white")
lines(x, sin(pi*x), col=pal[3], lwd=2)
lines(x, x^2-1, col=pal[9], lwd=2)
```

## Relationship of area and slope

Slope = $\dfrac{\int_0^\pi sin(x)dx}{\pi}=\dfrac{-cos(x)(-cos(0))}{\pi}=\dfrac{2}{\pi}$

```{r}
x <- seq(-2*pi, 2*pi, 0.1)
y <- sin(x)
y1 <- -cos(x)
y2 <- 2*x/pi

plot(x, y, type="l", col=pal[1], lwd=2, ylim=c(-3,3))
abline(h=0, v=0, lwd=2)
lines(x, y1, col=pal[3])
abline(v=pi, lty=2)
abline(h=-1, lty=2)
abline(a=-1, b=2/pi, lty=2)
```

Notice how the 3 dotted lines form a triangle. The curved orange area above and below the sloped line cancel each other out, such that the area from $\int_0^\pi sinx=\pi$

Why does the slope represent an *average* value of x in that region?

$Average=\dfrac{\int_a^b f(x) dx}{b-a}=\dfrac{F(b)-F(a)}{b-a}$


# Taylor Expansions 

Taylor expansions allow us to approximate solutions/functions when $x\rightarrow 0$. Largely about taking non-polynomial functions and finding polynomials that approximate them near some angle. Why do this? Polynomials are *much* easier to deal with than other functions, such as cosine, etc. They are easier to compute derivatives, integrals.

## General form

At $x-0$:

$P(x)=f(0) + \dfrac{df}{dx}(0)\dfrac{x^1}{1!} + \dfrac{d^2f}{dx^2}(0)\dfrac{x^2}{2!} + \dfrac{d^3f}{dx^3}(0)\dfrac{x^3}{3!} + \dfrac{d^4f}{dx^4}(0)\dfrac{x^4}{4!} + ...$

At any point $x=a$:

$P(x)=f(a) + \dfrac{df}{dx}(a)\dfrac{(x-a)^1}{1!} + \dfrac{d^2f}{dx^2}(a)\dfrac{(x-a)^2}{2!} + \dfrac{d^3f}{dx^3}(a)\dfrac{(x-a)^3}{3!} + \dfrac{d^4f}{dx^4}(a)\dfrac{(x-a)^4}{4!} + ...$

**Definitions**:

- **Taylor series**: the infinite sum, including *all* polynomial terms

- **Convergent series**: converges to the actual solution of the original function

- **Divergent series**: after reaching a certain threshold, will never converge and rather overshoot back and forth

  + **Radius of convergence**: the aforementioned threshold before which the series converges but after which the series diverges

## Example: $e^x$

The easiest example of a Taylor series is $e^x$, since $\dfrac{d(e^x)}{dx}=e^x$, i.e. its derivative is the same as the original function! So using the general form above...

$P(e^x)=1 + x + \dfrac{1}{2}x^2 + \dfrac{1}{6}x^3 + \dfrac{1}{24}x^4 + ...$

## Example: $e^{-x^2}$

Will need the *chain rule* for step (2)...

$\dfrac{d}{dx}(g(h(x))=\dfrac{dg}{dh}\dfrac{dh}{dx}=\dfrac{dg}{dx}$

Steps:

1. *Original*: $e^{0^2}=1$; *polynomial*: $P(x)=1 + ...$
2. *1st derivative*: $\dfrac{d(e^{-x^2})}{dx}= \dfrac{dg}{dh}\dfrac{dh}{dx}=\dfrac{dg}{dx}= ln(e)e^{h(x)}(-2x)=-e^{-x^2}2x=-(0)^2ln(0)=0$; *polynomial*: $P(x)=1 + 0x + ...$
3. *2nd derivative*: $\dfrac{-e^{-x^2}2x}{dx}=-2e^{-x^2} + 2xln(e)e^{-x2}==-2e^{-x^2} + 2xe^{-x2}=-2$; *polynomial*: $P(x)=1 + 0x + -2\dfrac{1}{2}x^2 +...=1 + 0x - 2x^2 +...$
4. ...

$P(e^{-x^2})=1 + 0x - 2(\dfrac{1}{2})x^2 + 0x^3 + \dfrac{1}{2}x^4 + ...$


```{r, fig.height=4, fig.width=8}
## first derivative
D( expression( e^(-x^2) ), 'x') # -(e^(-x^2) * (log(e) * (2 * x)))

e <- exp(1)

test_fun <- function(x) {
  e <- exp(1)
  y <- e^(-x^2)
  return(y)
}

tay_fun <- function(x) {
  e <- exp(1)
  y <- 1-(x^2)+0.5*x^4
  return(y)
}

test_fun2 <- function(x) {
  e <- exp(1)
  y <- e^x
  return(y)
}

tay_fun2 <- function(x) {
  e <- exp(1)
  y <- 1+x+(0.5*x^2)
  return(y)
}

tay_fun2_2 <- function(x) {
  e <- exp(1)
  y <- 1+x+(0.5*x^2)+((1/6)*(x^3))
  return(y)
}

tay_fun2_3 <- function(x) {
  e <- exp(1)
  y <- 1+x+(0.5*x^2)+((1/6)*(x^3))+((1/24)*(x^4))
  return(y)
}

x <- seq(-5, 5, by=0.1)

par(mfrow=c(1,2))

plot(x, test_fun2(x), type="l", col=cl1, lwd=3, ylim=c(-0.5,2))
lines(x, tay_fun2(x), col=pal[3], lwd=2)
lines(x, tay_fun2_2(x), col=pal[5], lwd=2)
lines(x, tay_fun2_3(x), col=pal[8], lwd=2)
abline(h=0, v=0, lwd=4)

plot(x, test_fun(x), type="l", col=cl1, lwd=3, ylim=c(-0.5,2))
lines(x, tay_fun(x), col=cl2, lwd=2)
abline(h=0, v=0, lwd=4)

```

## Example: $y=cosx$

Cosine form:

$cos(x)=1$ at $x=0$. Derivative is $\dfrac{d(cos)}{dx}(0)=-sin(0)=0$.

Polynomial form:

Generally, $P(x)=c_0 + c_1x + c_2x^2$. We know that $cos(0)=1$; therefore, $c_0$, giving us $P(x)=1 + c_1x + c_2x^2$. Deriving that expression, we get $\dfrac{dP}{dx}(0)=c_1 + 2c_20=c_1$.

$c_0=1$
$c_1=0$

Now we have $P(x)=1 + 0x + c_2x^2$ and are free to change $x_2$ so we can take advantage of the fact that the derivative $-sinx$ curves downward upon moving away from 0, i.e. it has a negative second derivative. Even though the rate of change is 0 at that point, the rate of change itself is decreasing at that point. <mark >Making sure that the second derivatives match will ensure that the original curve and polynomial curve at the same rate</mark>. 

$\dfrac{d^2P}{dx^2}(x)=2c_2=-1$
$c_2=-1/2$

Steps:

1. *Original*: $cos(x) \xrightarrow {x \rightarrow 0}1$; *polynomial*: $P(x)=1 +...$
  + 1st term is 1 b/c cos(0) is 1
  + 2nd term is 0 b/c -sin(0) is 0
2. *1st derivative*: $\dfrac{d(cosx)}{dx}=-sinx=-sin0=0$  $P(x)=1 + 0c...$
2. *2nd derivative*: $-sin(0)=0$; *polynomial*: $\dfrac{dP}{dx}(x)=0 + 0x + 2c_2x$
3. *2nd derivative*: $-cos(0)=-1$; *polynomial*: $\dfrac{d^2P}{dx^2}(x)=2c_2$

**Polynomial**: $P(x)=1 - \dfrac{1}{2}x^2$

```{r}
## first and second derivatives
D( expression( cos(x) ), "x" ) # -sinx
D( expression( -sin(x) ), "x" ) # -cosx

## set colors
pal <- brewer.pal(8, name="Spectral")
cl1 <- "red4"
cl2 <- "peru"

## initiate x
x <- seq(-10,10,by=0.2)

## plot
plot(x, cos(x), type="l", col=cl1, lwd=3, ylim=c(-pi,pi))
lines(x, -sin(x), col=cl2)
lines(x, -cos(x), col="wheat")
abline(h=0, v=0, lwd=2)
abline(v=pi, col="navy", lty=2)
abline(v=-pi, col="navy", lty=2)

# original solution is 1
abline(h=1, col="navy", lwd=2)

# 1st derivative solution is 0
abline(h=0, col="blue", lwd=2)

# 2nd derivative solution is -1
abline(h=-1, col="lightblue", lwd=2)
abline(a=0, b=-1, col="lightblue", lwd=2)

# taylor expansion
lines(x, (1-(0.5*x^2)), col="orange", lwd=2)
lines(x, (1-(0.5*x^2)+(1/6)*x^3), col="orange3", lwd=2) # add 3rd term, 1/6 b/c 3*2*1=6
lines(x, (1-(0.5*x^2)+(1/24)*x^4), col="orange4", lwd=2) # add 4th term, 1/24 b/c 4*3*2*1=24
```

Now let's try with brewer palette colors...

```{r}
## plot
n <- 11
pal <- brewer.pal(11, name="Spectral")
plot(x, cos(x), type="l", col=pal[[7]], lwd=3, ylim=c(-pi,pi))
lines(x, -sin(x), col=pal[[6]])
lines(x, -cos(x), col=pal[[5]])
abline(h=0, v=0, lwd=2)
abline(v=pi, lty=2)
abline(v=-pi, lty=2)

# original solution is 1
abline(h=1, col=pal[[n]], lwd=2)

# 1st derivative solution is 0
abline(h=0, col=pal[[n-1]], lwd=2)

# 2nd derivative solution is -1
abline(h=-1, col=pal[[n-2]], lwd=2)
abline(a=0, b=-1, col=pal[[n-2]], lwd=2)

# taylor expansion
lines(x, (1-(0.5*x^2)), col=pal[[1]], lwd=2)
lines(x, (1-(0.5*x^2)+(1/6)*x^3), col=pal[[2]], lwd=2) # add 3rd term, 1/6 b/c 3*2*1=6
lines(x, (1-(0.5*x^2)+(1/24)*x^4), col=pal[[3]], lwd=2) # add 4th term, 1/24 b/c 4*3*2*1=24

## plot
n <- 11
pal <- brewer.pal(n, name="BrBG")
plot(x, cos(x), type="l", col=pal[[7]], lwd=3, ylim=c(-pi,pi))
lines(x, -sin(x), col=pal[[6]])
lines(x, -cos(x), col=pal[[5]])
abline(h=0, v=0, lwd=2)
abline(v=pi, lty=2)
abline(v=-pi, lty=2)

# original solution is 1
abline(h=1, col=pal[[n]], lwd=2)

# 1st derivative solution is 0
abline(h=0, col=pal[[n-1]], lwd=2)

# 2nd derivative solution is -1
abline(h=-1, col=pal[[n-2]], lwd=2)
abline(a=0, b=-1, col=pal[[n-2]], lwd=2)

# taylor expansion
lines(x, (1-(0.5*x^2)), col=pal[[1]], lwd=2)
lines(x, (1-(0.5*x^2)+(1/6)*x^3), col=pal[[2]], lwd=2) # add 3rd term, 1/6 b/c 3*2*1=6
lines(x, (1-(0.5*x^2)+(1/24)*x^4), col=pal[[3]], lwd=2) # add 4th term, 1/24 b/c 4*3*2*1=24

```

So how good is this approximation? Let's try it out for, say, 0.1...

```{r}
cos(0.1)
1-0.5*(0.1^2)
```

Pretty damn close!


# Golden ratio

$x_{n+1}=1+\dfrac{1}{x_n}$

```{r}
rand_nums <- rnorm(20, mean=0, sd=50)

n_iter <- 20
mat <- matrix(data=NA, nrow = length(rand_nums), ncol=n_iter)
for(i in 1:length(rand_nums)) {
  x <- rand_nums[i]
  for(j in 1:n_iter) {
    ans_tmp <- 1+(1/x)
    mat[i,j] <- ans_tmp
    x <- ans_tmp
  }
}

mat
```

The golden ratio is the **fixed point**, $\phi$! No matter where you start, you end up there.

## Cobbweb plots

Ref.: https://bayesianbiologist.com/2012/07/13/dynamical-systems-mapping-chaos-with-r/

```{r}
q_map<-function(r=1,x_o=runif(1,0,1),N=100,burn_in=0,...)
{
  par(mfrow=c(2,1),mar=c(4,4,1,2),lwd=2)
  ############# Trace #############
  x<-array(dim=N)
  x[1]<-x_o
  for(i in 2:N)
    x[i]<-r*x[i-1]*(1-x[i-1])
  
  plot(x[(burn_in+1):N],type='l',xlab='t',ylab='x',...)
  #################################
  
  ##########  Quadradic Map ########
  x<-seq(from=0,to=1,length.out=100)
  x_np1<-array(dim=100)
  for(i in 1:length(x))
    x_np1[i]<-r*x[i]*(1-x[i])
  
  plot(x,x_np1,type='l',xlab=expression(x[t]),ylab=expression(x[t+1]))
  abline(0,1)
  
  start=x_o
  vert=FALSE
  lines(x=c(start,start),y=c(0,r*start*(1-start)) )
  for(i in 1:(2*N))
  {
    if(vert)
    {
      lines(x=c(start,start),y=c(start,r*start*(1-start)) )
      vert=FALSE
    }
    else
    {
      lines(x=c(start,
                r*start*(1-start)),
            y=c(r*start*(1-start),
                r*start*(1-start)) )
      vert=TRUE
      start=r*start*(1-start)
    }
  }
  #################################
}

q_map(r=3.84,x_o=0.4)
```

```{r}
q_map<-function(x_o=runif(1,0,1), N=20, burn_in=0,...)
{
  par(mfrow=c(1,2), lwd=2) #mar=c(4,4,1,2), 
  ############# Trace #############
  x <- array(dim=N)
  x[1] <- x_o
  for(i in 2:N)
    #x[i]<-r*x[i-1]*(1-x[i-1])
    x[i]<-1+(1/x[i-1])
  
  plot(x[(burn_in+1):N], type='l', xlab='t', ylab='x',...)
  #################################
  
  ##########  Quadradic Map ########
  x<-seq(from=0,to=5,length.out=100)
  x_np1<-array(dim=100)
  for(i in 1:length(x))
    #x_np1[i]<-r*x[i]*(1-x[i])
    x_np1[i] <- 1+(1/x[i])
    
  plot(x, x_np1, type='l', col="navy", lwd=3, xlab=expression(x[t]), ylab=expression(x[t+1]), xlim=c(-1,10), ylim=c(-5,5))
  abline(0,1, col="lightblue")
  abline(h=0, v=0, lwd=3)
  
  start=x_o
  vert=FALSE
  lines( x=c(start,start), y=c(0,1+1/start), col=pal[1] )
  for(i in 1:(2*N))
  {
    if(vert)
    {
      lines( x=c(start,start), y=c(start,1+1/start), col=pal[2] ) 
      vert=FALSE
    }
    else
    {
      lines(x=c(start,
                1+1/start),
            y=c(1+1/start,
                1+1/start), col=pal[3] )
      vert=TRUE
      start=1+1/start
    }
  }
  #################################
}

q_map(r_o=0.4)
```

## Bifurcation plot

```{r}
library(parallel)
bifurcation <- function(from=3,to=4,res=500,
                      x_o=runif(1,0,1),N=10,reps=10,cores=4)
{
  r_s<-seq(from=from,to=to,length.out=res)
  r<-numeric(res*reps)
  for(i in 1:res)
    r[((i-1)*reps+1):(i*reps)]<-r_s[i]
  
  x<-array(dim=N)
  
  iterate<-mclapply(1:(res*reps),
                    mc.cores=cores,
                    function(k){
                      x[1]<-runif(1,0,1)
                      for(i in 2:N)
                        x[i]<-r[k]*x[i-1]*(1-x[i-1])
                      
                      return(x[N])
                    })
  
  plot(r,iterate,pch=15,cex=0.1)
  
  return(cbind(r,iterate))
}

#warning: Even in parallel with 4 cores, this is by no means fast code!
bi<-bifurcation()
#png('chaos.png',width=1000,height=850)
#par(bg='black',col='green',col.main='green',cex=1)
#plot(bi,col='green',xlab='R',ylab='n --> inf',main='',pch=15,cex=0.2)
#dev.off()
```
