---
title: "Taylor Series"
author: "Emily Schwenger"
date: "`r Sys.Date()`"
output: 
  rmdformats::readthedown:
      self_contained: true
      thumbnails: false
      lightbox: true
      gallery: false
      highlight: tango
      df_print: kable
      fig_height: 6.5
      fig_width: 10
      css: custom.css
      toc_depth: 3
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE) 
```

# Setup

## Load libraries

```{r}
suppressPackageStartupMessages( {
  library(RColorBrewer)
} )
```

## Color palettes

```{r}
pal <- brewer.pal(11, name="Spectral") 
cl1 <- "red4"
cl2 <- "peru"
```

# Introduction 

Taylor expansions allow us to approximate solutions/functions when $x\rightarrow 0$. Largely about taking non-polynomial functions and finding polynomials that approximate them near some angle. Why do this? Polynomials are *much* easier to deal with than other functions, such as cosine, etc. They are easier to compute derivatives, integrals.

## General form

At $x-0$:

$P(x)=f(0) + \dfrac{df}{dx}(0)\dfrac{x^1}{1!} + \dfrac{d^2f}{dx^2}(0)\dfrac{x^2}{2!} + \dfrac{d^3f}{dx^3}(0)\dfrac{x^3}{3!} + \dfrac{d^4f}{dx^4}(0)\dfrac{x^4}{4!} + ...$

At any point $x=a$:

$P(x)=f(a) + \dfrac{df}{dx}(a)\dfrac{(x-a)^1}{1!} + \dfrac{d^2f}{dx^2}(a)\dfrac{(x-a)^2}{2!} + \dfrac{d^3f}{dx^3}(a)\dfrac{(x-a)^3}{3!} + \dfrac{d^4f}{dx^4}(a)\dfrac{(x-a)^4}{4!} + ...$

**Definitions**:

- **Taylor series**: the infinite sum, including *all* polynomial terms

- **Convergent series**: converges to the actual solution of the original function

- **Divergent series**: after reaching a certain threshold, will never converge and rather overshoot back and forth

  + **Radius of convergence**: the aforementioned threshold before which the series converges but after which the series diverges

# Practice

The easiest example of a Taylor series is $e^x$, since $\dfrac{d(e^x)}{dx}=e^x$, i.e. its derivative is the same as the original function! So using the general form above...

$P(e^x)=1 + x + \dfrac{1}{2}x^2 + \dfrac{1}{6}x^3 + \dfrac{1}{24}x^4 + ...$

For $e^{-x^2}$,

$P(e^{-x^2})=1 + -2(\dfrac{1}{2})x^2 + 0x^3 + \dfrac{1}{2}x^4 + ...$

1. $e^{0^2}=1$
2. ...

## Example: $e^{-x^2}$, $e^x$

```{r, fig.width=16}
e <- exp(1)

test_fun <- function(x) {
  e <- exp(1)
  y <- e^(-x^2)
  return(y)
}

tay_fun <- function(x) {
  e <- exp(1)
  y <- 1-(x^2)+0.5*x^4
  return(y)
}

test_fun2 <- function(x) {
  e <- exp(1)
  y <- e^x
  return(y)
}

tay_fun2 <- function(x) {
  e <- exp(1)
  y <- 1+x+(0.5*x^2)
  return(y)
}

tay_fun2_2 <- function(x) {
  e <- exp(1)
  y <- 1+x+(0.5*x^2)+((1/6)*(x^3))
  return(y)
}

tay_fun2_3 <- function(x) {
  e <- exp(1)
  y <- 1+x+(0.5*x^2)+((1/6)*(x^3))+((1/24)*(x^4))
  return(y)
}

x <- seq(-5, 5, by=0.1)

par(mfrow=c(1,2))

plot(x, test_fun2(x), type="l", col=cl1, lwd=3, ylim=c(-0.5,2))
lines(x, tay_fun2(x), col=pal[3], lwd=2)
lines(x, tay_fun2_2(x), col=pal[5], lwd=2)
lines(x, tay_fun2_3(x), col=pal[8], lwd=2)
abline(h=0, v=0, lwd=4)

plot(x, test_fun(x), type="l", col=cl1, lwd=3, ylim=c(-0.5,2))
lines(x, tay_fun(x), col=cl2, lwd=2)
abline(h=0, v=0, lwd=4)

```

## Example: $y=cosx$

Cosine form:

$cos(x)=1$ at $x=0$. Derivative is $\dfrac{d(cos)}{dx}(0)=-sin(0)=0$.

Polynomial form:

Generally, $P(x)=c_0 + c_1x + c_2x^2$. We know that $cos(0)=1$; therefore, $c_0$, giving us $P(x)=1 + c_1x + c_2x^2$. Deriving that expression, we get $\dfrac{dP}{dx}(0)=c_1 + 2c_20=c_1$.

$c_0=1$
$c_1=0$

Now we have $P(x)=1 + 0x + c_2x^2$ and are free to change $x_2$ so we can take advantage of the fact that the derivative $-sinx$ curves downward upon moving away from 0, i.e. it has a negative second derivative. Even though the rate of change is 0 at that point, the rate of change itself is decreasing at that point. <mark >Making sure that the second derivatives match will ensure that the original curve and polynomial curve at the same rate</mark>. 

$\dfrac{d^2P}{dx^2}(x)=2c_2=-1$
$c_2=-1/2$

Steps:

1. *Original*: $cos(x) \xrightarrow {x \rightarrow 0}1$; *polynomial*: $P(x)=1 + 0x + c_2x^2$
  + 1st term is 1 b/c cos(0) is 1
  + 2nd term is 0 b/c -sin(0) is 0
2. *1st derivative*: $-sin(0)=0$; *polynomial*: $\dfrac{dP}{dx}(x)=0 + 2c_2x$
3. *2nd derivative*: $-cos(0)=-1$; *polynomial*: $\dfrac{d^2P}{dx^2}(x)=2c_2$

**Polynomial**: $P(x)=1 - \dfrac{1}{2}x^2$

```{r}
## first and second derivatives
D( expression( cos(x) ), "x" ) # -sinx
D( expression( -sin(x) ), "x" ) # -cosx

## set colors
pal <- brewer.pal(8, name="Spectral")
cl1 <- "red4"
cl2 <- "peru"

## initiate x
x <- seq(-10,10,by=0.2)

## plot
plot(x, cos(x), type="l", col=cl1, lwd=3, ylim=c(-pi,pi))
lines(x, -sin(x), col=cl2)
lines(x, -cos(x), col="wheat")
abline(h=0, v=0, lwd=2)
abline(v=pi, col="navy", lty=2)
abline(v=-pi, col="navy", lty=2)

# original solution is 1
abline(h=1, col="navy", lwd=2)

# 1st derivative solution is 0
abline(h=0, col="blue", lwd=2)

# 2nd derivative solution is -1
abline(h=-1, col="lightblue", lwd=2)
abline(a=0, b=-1, col="lightblue", lwd=2)

# taylor expansion
lines(x, (1-(0.5*x^2)), col="orange", lwd=2)
lines(x, (1-(0.5*x^2)+(1/6)*x^3), col="orange3", lwd=2) # add 3rd term, 1/6 b/c 3*2*1=6
lines(x, (1-(0.5*x^2)+(1/24)*x^4), col="orange4", lwd=2) # add 4th term, 1/24 b/c 4*3*2*1=24
```

Now let's try with brewer palette colors...

```{r}
## plot
n <- 11
pal <- brewer.pal(11, name="Spectral")
plot(x, cos(x), type="l", col=pal[[7]], lwd=3, ylim=c(-pi,pi))
lines(x, -sin(x), col=pal[[6]])
lines(x, -cos(x), col=pal[[5]])
abline(h=0, v=0, lwd=2)
abline(v=pi, lty=2)
abline(v=-pi, lty=2)

# original solution is 1
abline(h=1, col=pal[[n]], lwd=2)

# 1st derivative solution is 0
abline(h=0, col=pal[[n-1]], lwd=2)

# 2nd derivative solution is -1
abline(h=-1, col=pal[[n-2]], lwd=2)
abline(a=0, b=-1, col=pal[[n-2]], lwd=2)

# taylor expansion
lines(x, (1-(0.5*x^2)), col=pal[[1]], lwd=2)
lines(x, (1-(0.5*x^2)+(1/6)*x^3), col=pal[[2]], lwd=2) # add 3rd term, 1/6 b/c 3*2*1=6
lines(x, (1-(0.5*x^2)+(1/24)*x^4), col=pal[[3]], lwd=2) # add 4th term, 1/24 b/c 4*3*2*1=24

## plot
n <- 11
pal <- brewer.pal(n, name="BrBG")
plot(x, cos(x), type="l", col=pal[[7]], lwd=3, ylim=c(-pi,pi))
lines(x, -sin(x), col=pal[[6]])
lines(x, -cos(x), col=pal[[5]])
abline(h=0, v=0, lwd=2)
abline(v=pi, lty=2)
abline(v=-pi, lty=2)

# original solution is 1
abline(h=1, col=pal[[n]], lwd=2)

# 1st derivative solution is 0
abline(h=0, col=pal[[n-1]], lwd=2)

# 2nd derivative solution is -1
abline(h=-1, col=pal[[n-2]], lwd=2)
abline(a=0, b=-1, col=pal[[n-2]], lwd=2)

# taylor expansion
lines(x, (1-(0.5*x^2)), col=pal[[1]], lwd=2)
lines(x, (1-(0.5*x^2)+(1/6)*x^3), col=pal[[2]], lwd=2) # add 3rd term, 1/6 b/c 3*2*1=6
lines(x, (1-(0.5*x^2)+(1/24)*x^4), col=pal[[3]], lwd=2) # add 4th term, 1/24 b/c 4*3*2*1=24

```

So how good is this approximation? Let's try it out for, say, 0.1...

```{r}
cos(0.1)
1-0.5*(0.1^2)
```

Pretty damn close!







